Mergeable Types

### 2017-02-28 paper/evaluation.tex

In this section, we present an evaluation of \name programming model
with the aim of assessing its practical utility for describing
geo-distributed eventually consistent applications. In particular, we
are interested in the horizontal scalability of applications built
using \name. Typical distributed eventually consistent data stores use
custom synchronization and dissemination protocols for transferring
updates between replicas. In contrast, \name is realized over Irmin~\cite{Irmin}
which uses Git transfer protocol~\cite{..}, a protocol not designed
with distributed data stores in mind. Hence, we also evaluate the
performance of synchronization and merging.

>>>>

In this section, we present an evaluation of \name programming model with the
aim of assessing its practical utility for describing geo-distributed
eventually consistent applications. In particular, we are interested in the
horizontal scalability of applications built using \name. Typical distributed
eventually consistent data stores use custom synchronization and dissemination
protocols for transferring updates between replicas. In contrast, \name is
realized over Irmin~\cite{irmin} which uses Git transfer protocol~\cite{..}, a
protocol not designed with distributed data stores in mind. Hence, we also
evaluate the performance of synchronization and merging.

### 2017-02-28 paper/intro.tex

To illustrate how these kinds of problems manifest in practice,
consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instantaneously witnessed on another, which may be in the process of
simultaneously attempting to perform its own \C{Add} or \C{Mult}
action.  While synchronizing the activities of all nodes to ensure at
most one such operation is performed at a time is impractical,
designating a single node to hold the counter state eschewing
replication altogether (as in a client-server
configuration~\cite{Armstrong}), is also an undesirable solution,
given the sensitivity of such architectures to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removing coordination altogether by simply
replicating counter state without having any supporting consistency
protocol is an equally infeasible approach.

@@

\vspace*{-.85in}

@@

directly in terms of the counter value maintained by each node.  In
doing so, we observe that each replica essentially operates over its
own version of a counter.  Local operations on a replicated object can
be thought of as yielding new versions, thus producing a version tree,
with one branch for each replica.  Thus, every branch represents
different (immutable) versions maintained by different nodes, with the
state produced by the computation performed over a counter on a node
recorded along the node's local branch for the counter.  Now, to
generate a globally consistent view of a counter, we only need to
define a merge operation that explains how to combine two local
versions to produce a new version that reflects both their states.

@@

In this paper, we develop a programming model that realizes a monadic
version control system centered around data, rather than file,
coherence.  The \name monad lets programmers write and compose
concurrent computations around multiple (implicit) versions of a
mergeable datatype, an ordinary ML datatype additionally equipped with
a \emph{merge} function responsible for deriving a consistent global
state from a collection of local versions of that state.  A
computation progresses by forking (i.e., replicating) new versions of
existing versions on geo-distributed nodes, allowing
synchronization-free local computation to proceed on those nodes,
creating new versions along an existing \emph{branch} that represents
updated local instances, and by merging branches to realize global
consistency.

Our model allows ML programmers to get the benefits of achieving
highly available (low-latency) distributed computation, while
continuing to enjoy the comforts of high-level reasoning and the
familiarity of using standard libraries already provided by the
language.  Notably, while \name's programming model makes no explicit
reference to any specific operational manifestation of a distributed
system (e.g., programmers do not need to explicitly manage replicas),
we demonstrate that it can be nonetheless efficiently realized on
existing geo-replicated distributed systems.

@@

The remainder of the paper is organized as follows.  We present the
\name\ programming model informally in the next section.
Sec.~\ref{sec:mergeable_types} develops a series of examples that
illustrate how merge operations can be defined and used.
Sec.~\ref{sec:formalization} defines an operational semantics, and
formalizes our intuition of a mergeable type in terms of morphisms over
datatype operations.  We also present soundness (consistency) and
progress guarantees enjoyed by the
semantics. Sec.~\ref{sec:system-model} considers an instantiation of
the model suitable for distributed environments with network
partitions and failures.  Implementation details describing how we
incorporate mergeable types into OCaml are presented in
Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given in
Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
and~\ref{sec:conclusions} describes related work and conclusions.

>>>>

In this paper, we consider how declarative abstractions can be used to
overcome this tension to provide the best of both worlds: functional
programs whose logical structure can be seamlessly transplanted to a
distributed setting, while nonetheless being resilient to system
realities such as network failure and latency.  Our key insight is the
development of a principled methodology that allows us transform
low-level operational details such as replication and consistency
into high-level compositional reasoning on language-level datatypes.

To illustrate how the kinds of problems described above manifest in
practice, consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instantaneously witnessed on another, which may be in the
process of simultaneously attempting to perform its own \C{Add} or
\C{Mult} action.  While synchronizing the activities of all nodes to
ensure at most one such operation is performed at a time is
impractical, designating a single node to hold the counter state
eschewing replication altogether (as in a client-server
configuration~\cite{Armstrong}), is also an undesirable solution,
given the sensitivity of such architectures to network partitions and
server failures, and the negative performance impact it incurs in
geo-distributed environments~\cite{Walter}.  Removing coordination
altogether by simply replicating counter state without having any
supporting consistency protocol is an equally infeasible approach.

@@

\vspace*{-1.2in}

@@

directly in terms of the counter value maintained by each node?  To do
so, we first observe that each replica essentially operates over its
own version of a counter.  Thus, local operations on a replicated
object can be thought of as yielding new versions, collectively
producing a version tree, with one branch for each replica.  Every
branch represents different (immutable) versions maintained by
different nodes, with the state produced by the computation performed
over a counter on a node recorded along the node's local branch for
the counter.  Now, to generate a globally consistent view of a
counter, we only need to define a merge operation that explains how to
combine two local versions to produce a new version that reflects both
their states.  This operation is defined not in terms of replicas or
other system-specific artifacts, but in terms of the semantics of the
datatype itself.

@@

Our main innovation is thus the development of a programming model
that realizes a monadic version control system centered around data,
rather than file, coherence.  The \name monad lets programmers write
and compose concurrent computations around multiple (implicit)
versions of a mergeable datatype, an ordinary ML datatype additionally
equipped with a \emph{merge} function responsible for deriving a
consistent global state from a collection of local versions of that
state.  A computation progresses by forking (i.e., replicating) new
versions of existing versions, allowing synchronization-free local
computation to proceed on those nodes, creating new versions along an
existing \emph{branch} that represents updated local instances, and by
merging branches to realize global consistency.

Our model allows ML programmers to get the benefits of achieving
highly available (low-latency) distributed computation, while
continuing to enjoy the comforts of high-level reasoning and the
familiarity of using standard libraries already provided by the
language.  Notably, while \name's programming model makes no explicit
reference to any specific operational manifestation of a distributed
system (e.g., programmers do not need to explicitly manage replicas),
we demonstrate that it can be nonetheless efficiently realized on
existing real-world geo-replicated distributed systems.

@@

\noindent The remainder of the paper is organized as follows.  We present the
\name\ programming model informally in the next section.
Sec.~\ref{sec:mergeable_types} develops a series of examples that
illustrate how merge operations can be defined and used.
Sec.~\ref{sec:formalization} defines an operational semantics, and
formalizes our intuition of a mergeable type in terms of morphisms over
datatype operations.  We also present soundness (consistency) and
progress guarantees enjoyed by the
semantics. Sec.~\ref{sec:system-model} considers an instantiation of
the model suitable for distributed environments with network
partitions and failures.  Implementation details describing how we
incorporate mergeable types into OCaml are presented in
Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given in
Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
and~\ref{sec:conclusions} describes related work and conclusions.

### 2017-02-28 paper/model.tex

In this section, we describe the \name programming model through the
example of a collaborative drawing application we call \drawsome.
Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
application. \drawsome represents a free-hand drawing canvas in terms
of a tree of quadrants.  Each quadrant is simply a leaf node
containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
quadrant contains multiple pixels of different colors. Quadrants are
expanded into a tree structures as and when pixels are colored.  The
representation is thus optimized for sparse canvases, such as
whiteboards. The application supports three simple operations:
creating a new canvas, setting the pixel at a specified coordinate,
and returning the pixel at a given coordinate. The \C{merge} function
is explained below.  \drawsome lets multiple users collaborate on a
canvas that is conceptually shared among them. Under a shared-memory
abstraction, there would be a single copy of the canvas that is
updated concurrently by multiple clients; from the perspective of any
single client, the canvas could change without any explicit
intervention. \name ascribes functional semantics to sharing by
letting each client work on its own version of the state (the tree
data structure in this example), later merging concurrent versions
on-demand.  Thus, the primary artifact of the \name programming model
is a versioned data structure in which different versions are managed
by different clients.

@@

\name requires a three-way \C{merge} function to merge concurrent
versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}) that
includes includes two concurrent versions (\C{v1} and \C{v2}),
and their least common ancestor (\C{lca}) - the version from which the
two concurrent versions evolved independently. The merge function can
\begin{wrapfigure}{L}{.4\textwidth}
\centering
\includegraphics[scale=0.43]{Figures/canvas-merging}
\caption{Merging concurrent versions (\C{v1} and \C{v2}) of a drawing
canvas. \C{lca} is their common ancestor.}
\label{fig:canvas-merging}
\end{wrapfigure}
make use of the pixel values of the common ancestor to merge the pixel
values on both the canvases. For instance, if the color of a pixel in
\C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
is white, then it means that only \C{v2} modified the color. Hence the
pixel is colored green in the merged canvas. On the other hand, if the
pixel is red in \C{v1}, then it means that both \C{v1} and \C{v2} have
modified the color. In such case, an appopriate color-mixing algorithm
can be used to determine the color of pixel.  For instance, the pixel
can be colored yellow - an additive combination of red and green. The
logic is illustrated in Fig.~\ref{fig:canvas-merging}.
\begin{figure}
\begin{center}
  \begin{ocaml}
let color_mix px1 px2 : pixel =
let f = Char.code in
let h x y = Char.chr @@ (x + y)/ 2 in
let (r1,g1,b1) = (f px1.r, f px1.g, f px1.b) in
let (r2,g2,b2) = (f px2.r, f px2.g, f px2.b) in
let (r,g,b) = (h r1 r2, h g1 g2, h b1 b2) in {r=r; g=g; b=b}

>>>>

In this section, we describe the \name programming model through the
example of a collaborative drawing application we call \drawsome.
Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
application. \drawsome represents a free-hand drawing canvas in terms
of a tree of quadrants.  Each quadrant is simply a leaf node
containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
quadrant contains multiple pixels of different colors. Quadrants are
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.43]{Figures/canvas-merging}
\caption{Merging concurrent versions (\C{v1} and \C{v2}) of a drawing
canvas. \C{lca} is their common ancestor.}
\label{fig:canvas-merging}
\end{wrapfigure}
expanded into a tree structures as and when pixels are colored.  The
representation is thus optimized for sparse canvases, such as
whiteboards. The application supports three simple operations:
creating a new canvas, setting the pixel at a specified coordinate,
and returning the pixel at a given coordinate. The \C{merge} function
is explained below.  \drawsome lets multiple users collaborate on a
canvas that is conceptually shared among them. Under a shared-memory
abstraction, there would be a single copy of the canvas that is
updated concurrently by multiple clients; from the perspective of any
single client, the canvas could change without any explicit
intervention. \name ascribes functional semantics to sharing by
letting each client work on its own version of the state (the tree
data structure in this example), later merging concurrent versions
on-demand.  Thus, the primary artifact of the \name programming model
is a versioned data structure in which different versions are managed
by different clients.

@@

\name requires a three-way \C{merge} function to merge concurrent
versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}) that
includes includes two concurrent versions (\C{v1} and \C{v2}),
and their least common ancestor (\C{lca}) - the version from which the
two concurrent versions evolved independently. The merge function can
make use of the pixel values of the common ancestor to merge the pixel
values on both the canvases. For instance, if the color of a pixel in
\C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
is white, then it means that only \C{v2} modified the color. Hence the
pixel is colored green in the merged canvas. On the other hand, if the
pixel is red in \C{v1}, then it means that both \C{v1} and \C{v2} have
modified the color. In such case, an appropriate color-mixing algorithm
can be used to determine the color of pixel.  For instance, the pixel
can be colored yellow - an additive combination of red and green. The
logic is illustrated in Fig.~\ref{fig:canvas-merging}.
\begin{figure}
\begin{center}
  \begin{ocaml}
let color_mix px1 px2 : pixel =
let f = Char.code in
let h x y = Char.chr @@ (x + y)/ 2 in
let (r1,g1,b1) = (f px1.r, f px1.g, f px1.b) in
let (r2,g2,b2) = (f px2.r, f px2.g, f px2.b) in
let (r,g,b) = (h r1 r2, h g1 g2, h b1 b2) in {r=r; g=g; b=b}

### 2017-02-28 paper/related.tex

Our idea of versioning state bears resemblance to Concurrent
Revisions~\cite{BBL+10}, a programming abstraction that provides
deterministic concurrent execution by forking and merging
\emph{versions}, portions of state shared among concurrent threads.
The idea of using revisions as a means to programming eventually
consistent distributed systems was further developed
in~\cite{BFL+12,Burckhardt2012}.  The \name\ programming model,
however, differs substantially from a concurrent revisions model
because it imposes no distinction between \emph{servers}, machines
that hold global state, and \emph{clients}, devices that operate on
local, potentially stale, data - any computation executing in a
distributed environment is free to fork new versions, and synchronize
against other replicated state.  Our model, furthermore, supports
fully decentralized operation and is robust to network partitions and
failures. Just as significantly, \name\ allows applications to
customize join semantics with programmable merge operations.  Indeed,
the integration of a version-based mechanism within OCaml allows a
degree of type safety, composability, and profitable use of
polymorphism not available in related systems.

\cite{Burckhardt2015} also presents an operational model of a
replicated data store that is based on the abstract system model
presented in ~\cite{Burckhardt2014}; their design is similar to the
model described in~\cite{pldi15}.  In both approaches, coordination
among replicas involves transmitting operations on replicated objects
that are performed locally on each replica.  In contrast,
\name\ allows programmers to use familiar state-based and functional
abstractions when developing distributed applications.  As we
illustrated in Fig.~\ref{fig:counter-rdt}, using effects and
operations to coordinate the activities of replicas may involve
addressing subtleties that do not manifest otherwise.  Our case
studies and experimental results support our contention that using
well-understood state (heap)-based abstractions to build distributed
applications greatly simplifies program reasoning and eases
development, without compromising efficiency.

@@

There have been numerous proposals over the years for realizing
distributed functionality in a functional programming context.
Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} both extend Standard ML
with distributed programming abstractions.  Kali-Scheme~\cite{CJK95}
defines a distributed extension for Scheme-48, and Cloud
Haskell~\cite{EBPJ11} describes a domain-specific language that
enables development of distributed Haskell programs.
Obliq~\cite{Car95} explores issues related to mobility, lexical
scoping, and network references in an object-oriented framework that
enables safe (i.e., lexical) transmission of closures.  However,
unlike \name, none of these systems consider issues of object
replication or consistency as central to their programming models.
The realities of modern-day scalable distributed systems, especially
those that are constructed from geographically distributed
participants, dictate that we must necessarily grapple with
consistency issues to achieve any kind of reasonable performance; it
is noteworthy that replication is inherent in many widely-used
distributed system implementations today (e.g.,
Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
only weak consistency guarantees.  Rather than providing, as other
approaches have, a low-level operational treatment of replication, our
primary contribution is the development of a principled high-level
approach to framing issues of consistency and replication.  Our goal
is to leverage functional programming principles to minimize
disruption to the way programmers structure and reason about their
applications, without compromising efficiency or performance.  Indeed,
the semantics of mergeable types and versioning has utility in any
concurrent setting that must deal with non-trivial coordination and
synchronization costs, even without taking distribution or replication
issues into account.

>>>>

Our idea of versioning state bears resemblance to Concurrent
Revisions~\cite{BBL+10}, a programming abstraction that provides
deterministic concurrent execution.  The idea of using revisions as a
means to programming eventually consistent distributed systems was
further developed in~\cite{BFL+12,BLF+12}.  The
\name\ programming model, however, differs substantially from a
concurrent revisions model because it imposes no distinction between
\emph{servers}, machines that hold global state, and \emph{clients},
devices that operate on local, potentially stale, data - any
computation executing in a distributed environment is free to fork new
versions, and synchronize against other replicated state.  Our model,
furthermore, supports fully decentralized operation and is robust to
network partitions and failures. Just as significantly, \name\ allows
applications to customize join semantics with programmable merge
operations.  Indeed, the integration of a version-based mechanism
within OCaml allows a degree of type safety, composability, and
profitable use of polymorphism not available in related systems.

\cite{Burckhardt2015} also presents an operational model of a
replicated data store that is based on the abstract system model
presented in ~\cite{Burckhardt2014}; their design is similar to the
model described in~\cite{pldi15}.  In both approaches, coordination
among replicas involves transmitting operations on replicated objects
that are performed locally on each replica.  In contrast,
\name\ allows programmers to use familiar state-based and functional
abstractions when developing distributed applications.  As we
illustrated in Fig.~\ref{fig:counter-rdt}, using effects and
operations to coordinate the activities of replicas may involve
addressing subtleties that would not manifest in their absence.  Our
case studies and experimental results support our contention that
using well-understood state (heap)-based abstractions to build
distributed applications greatly simplifies program reasoning and
eases development, without compromising efficiency.

@@

There have been numerous proposals over the years for realizing
distributed functionality in a functional programming context.
Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} extend Standard ML,
while Acute~\cite{SLW+05} and HashCaml~\cite{BSS+06} extend OCaml, with
distributed programming abstractions and support, with particular
focus on type-safe marshalling and mobility.  Kali-Scheme~\cite{CJK95}
defines a distributed extension for Scheme-48, and Cloud
Haskell~\cite{EBPJ11} describes a domain-specific language that
enables development of distributed Haskell programs.
Obliq~\cite{Car95} explores issues related to mobility, lexical
scoping, and network references in an object-oriented framework that
enables safe (i.e., lexical) transmission of closures.  However,
unlike \name, none of these systems consider issues of object
replication or consistency as central to their programming models.
The realities of modern-day scalable distributed systems, especially
those that are constructed from geographically distributed
participants, dictate that we must necessarily grapple with
consistency issues to achieve any kind of reasonable performance; it
is noteworthy that replication is inherent in many widely-used
cloud-based distributed system implementations today (e.g.,
Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
only weak consistency guarantees.  Rather than exposing a low-level
operational treatment of replication, foisting the burden of reasoning
about replicated state onto the programmer, our primary contribution
is the development of a principled high-level approach to framing
these issues that abstracts these kinds of system-level details.  Our
goal is to leverage functional programming principles to minimize
disruption to the way programmers structure and reason about their
applications, without compromising efficiency or performance.  Indeed,
the semantics of mergeable types and versioning has utility in any
concurrent setting that must deal with non-trivial coordination and
synchronization costs, even without taking distribution or replication
issues into account.
